\Title

### 2023-06-08 Grimaldi-etal-BiolCybernetics.tex
Update on Overleaf.

The present paper proposes a real-world application that extends a recent model of spiking neurons with heterogeneous synaptic delays~\cite{grimaldi_learning_2022}. Indeed, this previous model was trained to solve a simplified motion detection task on a synthetic event-based dataset generated by moving parameterized textures. This was a first demonstration that formal neurons can exploit the precise timing of spikes to improve neural computation thanks to heterogeneous delays. Here, we extend these results to a more complex and natural setting. First, we define the ecological cognitive task that the model must solve with the different datasets on which it will be tested. Instead of the synthetic textures that have been used to generate motion-driven stimuli, we use natural images and random movements that mimic biological saccades. We then develop the Heterogeneous Delays Spiking Neural Network (HD-SNN) model thanks to theoretically defined first principles and derive a learning rule to adapt heterogeneous synaptic delays. Using this detection task, we study the emergence of spatiotemporal spiking motifs when this single layer of spiking neurons is trained in a supervised manner. We investigate the efficiency of the motion detection mechanism and, in particular, its resilience to synaptic weight pruning. Indeed, once trained, the amount of event-driven computations could be drastically reduced by removing weak synapses while maintaining peak performance for the classification task. In this way, we will be able to show how such a model can provide an efficient solution to the energy/accuracy tradeoff.

@@

In this paper, we test whether the HD-SNN model is able to efficiently learn a motion detection task defined by a realistic event-based data stream. This type of signal is commonly captured by a Dynamic Vision Sensor (DVS) and is inspired by the signal sent from the retinal ganglion cells through the optic nerve to the visual cortex. The events are binary in nature and should be sufficient to perform the motion detection task in a fast and efficient manner. Here we first describe the task definition and then the HD-SNN model we use for motion inference and how we train it.

@@

To train and test our model, we need to define a visual dataset for which we supervise the direction and speed of motion. To achieve this, we define a procedure for animating a natural visual scene with virtual eye movements, similar to those used in neurobiological~\citep{vinje_sparse_2000, baudot_animation_2013} and computational neuroscience~\citep{kremkow_push-pull_2016} studies. First, we draw a trajectory inspired by the biological movements of the eyes. Indeed, these movements allow us to dynamically actuate the center of vision, or gaze, in the visual field. In animals with a fovea, this is particularly useful because it allows the area with the highest density of photoreceptors in the environment to be moved, for example, to a point of interest. A specific mechanism for this function is saccades, which are rapid eye movements that reposition the center of vision. In humans, saccades are very frequent (on average $2$ per second~\citep{dandekar_neural_2012}), very fast (about $80~\ms$), and have a wide range of speeds. On a more microscopic scale, the human gaze moves with a continuous drift similar to a Brownian trajectory~\citep{poletti_head-eye_2015}. To preserve the full generality of the task, we define eye movements using such a form of random walk~\citep{engbert_integrated_2011}. This approach first defines a finite set of possible 2D movements in polar coordinates. Based on the distribution of biological movements, we simplified it by selecting a set of eye movements as the Cartesian product of $XXX8$ linearly spaced directions and $XXX3$ geometrically spaced speeds. Next, we define a gaze trajectory as segments whose duration is drawn from a Poisson distribution with a mean block length of $24~\ms$, similar to a Lévy flight~\citep[p. 289]{mandelbrot_fractal_1982}. Finally, assuming first that the speeds are sampled uniformly and independently from the set of different motions, and second that the motion is uniform during a time segment,  the trajectory is generated by integrating the motion sequence. This generative model yield trajectories that are qualitatively very similar to those observed for human eye movements (\seeFig{motion_task}-\textit{(Left)}).

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \caption{
        {\bf - Motion Detection Task.} \textit{(Left)}~We use large natural images ($1024\times1024$) in which an aperture ($128\times128$) extracts a cropped image around the view axis. To mimic the effect of a saccadic eye movement, this axis moves according to a jagged random walk. We show such a trajectory with a length of $128$ time steps (green line). \textit{(Right)}~Snapshots of the synthetic event stream at different time steps (start and end frames marked by a white and black dot, respectively). The dynamics of the cropped image translated according to the trajectory as a function of time produces a naturalistic movie, which is then transformed into an event-based representation. Mimicking the retina, this representation encodes in each pixel of the image quantal increases or decreases in luminance, i.e. ON (red) and OFF (blue) events. In the lower left corner of the snapshots, the translation vector is shown in red as one of the possible classes of motion. Note the change in the direction of motion between the third and fourth image, and also that, due to the aperture problem, contours parallel to the motion emit relatively fewer spikes.
        }
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of $600$ natural images that were previously used to study the statistics of natural images~\citep{perrinet_edge_2015}. Note that these were pre-processed to be in grayscale and to equalize (i.e., whiten) the energy in each frequency band. This process is known to occur in the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $256 \times 256$ in size, and we will extract subimages of size $64 \times 64$ positioned around the center of gaze at each time step. We will discretize time in $1~\ms$ bins and produce movies of duration $N_T = 128~\ms$. To avoid boundary effects, we will randomly position the full trajectory in image space so that the subimage is translated using the position given by the trajectory at each time step without touching the boundaries. At each time step, the translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a subpixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the unit speed to correspond to a movement of one pixel per frame (i.e., per time bin).

To transform each movie into events, as recorded by a DVS, we compute a gradient image (initialized at zero) by computing the temporal gradient of the pixels' intensity over two successive frames. For a given pixel and time stamp, an event is generated when the absolute value of this gradient exceeds a threshold. The event has either an OFF or ON polarity, depending on whether the gradient is negative or positive. The signed threshold is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, i.e. a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF), and their time $\timev_\arank$ (\seeFig{motion_task}-\textit{(Right)}). The goal here is to infer the correct motion solely by observing these events.%%%-----------------------------------------------------------------
\subsection{The Heterogeneous Delays SNN (HD-SNN) model}
%----------------------------%
%
\begin{figure}%[t!]
    \centering
    \includegraphics[width=0.980\linewidth]{figures/HDSNN_graph.pdf}
    \caption{\textbf{The core mechanism of the HD SNN model.} \textit{(Top-Left)}~Two spiking motifs are emitted from four presynaptic neurons. Once integrated by the synapses of the postsynaptic neuron~\textit{(Right)}, the spiking motifs are shifted in time by the synaptic delays and weighted by the synaptic weights~\textit{(Middle-Left)}. When they reach the soma of the postsynaptic neuron, the different spikes contribute to a modification of its membrane potential according to an activation function. In this example, we use the same activation function as for a Leaky Integrate and Fire neuron~\textit{(Bottom-Left)}. The first spiking motif is synchronized by the synaptic delays and causes a sudden rise in the membrane potential of the postsynaptic neuron. An output spike is emitted when the membrane potential reaches the threshold and it is then reset.   \textit{(Right)}~An illustration of a spiking neuron with different synaptic weights (represented by the thickness of the synapses) and different synaptic delays (represented by the length of the synapses). }%
    \label{fig:izhikevich}%
\end{figure}%
% 
%----------------------------%
In a neurobiological recording or in the sensory signal obtained from an event-based camera, the input consists of a stream of \emph{spikes} or \emph{events}. This can be formalized as a list of tuples, each tuple representing neural addresses and timestamps. We defined this list as $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N} $ is the total number of events in the data stream and the rank $\arank$ is the index of each event in the list of events (\seeFig{izhikevich}-\textit{(Top-Left} for an illustration). Events are usually ordered by their time of occurrence. Each event has a time of occurrence $\timev_\arank$ and an associated address $\presynaddr_\arank$. This defines an address space $\presynaddrspace$, which consists of the set of possible addresses. In a neurobiological recording, this may be the identified set of all presynaptic neurons. For neuromorphic hardware, this can be defined as $[1, \Npol] \times [1, \Nx]\times[1, \Ny] \subset \mathbb{N}^3$, where $\Npol$ is the number of polarities ($\Npol=2$ for the ON and OFF polarities encoded in event-based cameras) and $(\Nx, \Ny)$ is the height and width of the image in pixels. Thus, an address $\presynaddr_\arank$ is typically in the form $(\polev_\arank, x_\arank, y_\arank)$ for event-based cameras.

@@

After training our model, we first analyze the weights learned for the different neurons (\seeFig{kernels}). We first notice that these cells are very localized along each trajectory, but show an orientation selectivity similar to that found for MT neurons~\citep{deangelis_functional_1999}. Focusing on the positive weights, a strong selectivity along specific motion axes is observed for each of the different kernels. These directions can be easily associated with the direction of motion controlled in the natural images and show a typical anisotropic profile~\citep{kaplan_anisotropic_2013}. For example, the first kernel shows a strong selectivity for a downward direction of motion. Note that as the delays become larger, the coefficients become lower and less localized, consistent with the average diffusion of information included in the generative model and with the diffusion introduced in motion-based prediction models~\citep{perrinet_motion-based_2012,khoei_flash-lag_2017}. We also observed a strong dependence between the weights reaching the ON polarities and those reaching the OFF polarities. In particular, whenever a weight for a given position and delay is positive for one polarity, it will be negative for the other. This property is due to the way the events are generated and the fact that the luminance cannot increase and decrease at the same time.  Interestingly, the relative organization of the receptive fields in quadrature of phase follows a push-pull organization predicted by~\citet{kremkow_push-pull_2016} to explain neurophysiological results obtained after showing similar natural scenes with synthetic eye movements~\citep{baudot_animation_2013}.

>>>>

The present paper proposes a real-world application that extends a recent model of spiking neurons with heterogeneous synaptic delays~\cite{grimaldi_learning_2022}. Indeed, this previous model was trained to solve a simplified motion detection task on a synthetic event-based dataset generated by moving parameterized textures. This was a first demonstration that formal neurons can exploit the precise timing of spikes to improve neural computation thanks to heterogeneous delays. Here, we extend these results to a more complex and natural setting. First, we define the ecological cognitive task that the model must solve with the different datasets on which it will be tested. Instead of the synthetic textures that had been used to generate motion-driven stimuli, we here use natural images and random-walk movements that mimic biological saccades. We then develop the Heterogeneous Delays Spiking Neural Network (HD-SNN) model from theoretically defined first principles and derive a learning rule to adapt heterogeneous synaptic delays. Using this detection task, we study the emergence of spatiotemporal spiking motifs when this single layer of spiking neurons is trained in a supervised manner. We investigate the efficiency of the motion detection mechanism and, in particular, its resilience to synaptic weight pruning. Indeed, once trained, the amount of event-driven computations could be drastically reduced by removing weak synapses while maintaining peak performance for the classification task. In this way, we will be able to show how such a model can provide an efficient solution to the energy/accuracy tradeoff.

@@

In this paper, we test whether the HD-SNN model is able to efficiently learn to solve a motion detection task defined by a realistic event-based data stream. This type of signal is commonly captured by a Dynamic Vision Sensor (DVS) and is inspired by the signal sent from the retinal ganglion cells through the optic nerve to the visual cortex. The events are binary in nature and should be sufficient to perform the motion detection task in a fast and efficient manner. Here we first describe the task definition and then the HD-SNN model we use for motion inference and how we train it.

@@

To train and validate our model which uses supervised learning, we need to define a visual dataset for which we explictly know the ground truth direction and speed of motion. To achieve this, we define a procedure for animating a natural visual scene with virtual eye movements, similar to those used in neurobiological~\citep{vinje_sparse_2000, baudot_animation_2013} and computational neuroscience~\citep{kremkow_push-pull_2016} studies. First, we draw a trajectory inspired by the biological movements of the eyes. Indeed, these movements allow us to dynamically actuate the center of vision, or gaze, in the visual field. In animals with a fovea, this is particularly useful because it allows the area with the highest density of photoreceptors in the environment to be moved, for example, to a point of interest. A specific mechanism for this function is saccades, which are rapid eye movements that reposition the center of vision. In humans, saccades are very frequent (on average $2$ per second~\citep{dandekar_neural_2012}), very fast (about $80~\ms$), and have a wide range of speeds. On a more microscopic scale, the human gaze moves with a continuous drift similar to a Brownian trajectory~\citep{poletti_head-eye_2015}. To preserve the full generality of the task, we define eye movements using such a form of random walk~\citep{engbert_integrated_2011}. This approach first defines a finite set of possible 2D movements in polar coordinates. Based on the distribution of biological movements, we simplified it by selecting a set of eye movements as the Cartesian product of $XXX8$ linearly spaced directions and $XXX3$ geometrically spaced speeds. Next, we define a gaze trajectory as segments whose duration is drawn from a Poisson distribution with a mean block length of $24~\ms$, similar to a Lévy flight~\citep[p. 289]{mandelbrot_fractal_1982}. Finally, assuming first that the speeds are sampled uniformly and independently from the set of different motions, and second that the motion is uniform during a time segment,  the trajectory is generated by integrating the motion sequence. This generative model yield trajectories that are qualitatively very similar to those observed for human eye movements (\seeFig{motion_task}-\textit{(Left)}).

%%%-----------------------------------------------------------------
%: fig:motion_task
%%%-----------------------------------------------------------------
\begin{figure}%[h!]%[ht!]
    \centering
    \includegraphics[width=0.95\linewidth]{figures/motion_task.pdf}
    \caption{
        {\bf - Motion Detection Task.} \textit{(Left)}~We use large natural images ($1024\times1024$) in which an aperture ($128\times128$) extracts a cropped image around the view axis. To mimic the effect of a saccadic eye movement, this axis moves according to a jagged random walk. We show such a trajectory with a length of $256$ time steps (green line). \textit{(Right)}~Snapshots of the synthetic event stream at different time steps (start and end frames marked by a white and black dot, respectively). The dynamics of the cropped image translated according to the trajectory as a function of time produces a naturalistic movie, which is then transformed into an event-based representation. Mimicking the retina, this representation encodes in each pixel of the image quantal increases or decreases in luminance, i.e. ON (red) and OFF (blue) events. In the lower left corner of the snapshots, the translation vector is shown in red as one of the possible classes of motion. Note the change in the direction of motion between the third and fourth image, and also that, due to the aperture problem, contours parallel to the motion emit relatively fewer spikes.
        }
    \label{fig:motion_task}
\end{figure}
%%%-----------------------------------------------------------------
Once these eye trajectories are generated, we can apply them to a visual scene. For this purpose, we selected a database of $600$ natural images that were previously used to study the statistics of natural images~\citep{perrinet_edge_2015}. Note that these were pre-processed to be in grayscale and to equalize (i.e., whiten) the energy in each frequency band. This process is known to occur in the retino-thalamic pathway~\citep{dan_efficient_1996}. These images are $1024 \times 1024$ in size, and we will extract subimages of size $128 \times 128$ positioned around the center of gaze at each time step. We will discretize time in $1~\ms$ bins and produce movies of duration $N_T = 128~\ms$. To avoid boundary effects, we will randomly position the full trajectory in image space so that the subimage is translated using the position given by the trajectory at each time step without touching the boundaries. At each time step, the translation is computed using a coordinate roll in the horizontal and vertical dimensions, followed by a subpixel translation defined in Fourier space~\citep{perrinet_sparse_2015}. Note that the magnitude of the displacement is relative to the time bin, and we have defined the unit speed to correspond to a movement of one pixel per frame (i.e., per time bin).

To transform each movie into events, as recorded by a DVS, we compute a gradient image (initialized at zero) by computing the temporal gradient of the pixels' intensity over two successive frames. For a given pixel and time stamp, an event is generated when the absolute value of this gradient exceeds a threshold. The event has either an OFF or ON polarity, depending on whether the gradient is negative or positive. The signed threshold is then subtracted from the residual gradient image. When applied to the whole movie, the event stream is similar to the output of a neuromorphic camera~\citep{rasetto_challenges_2022}, i.e. a list of events defined by $x_\arank$ and $y_\arank$ (their position on the pixel grid), their polarity $\polev_\arank$ (ON or OFF), and their time $\timev_\arank$ (\seeFig{motion_task}-\textit{(Right)}). The goal here is to infer the correct motion solely by observing these events.%%%-----------------------------------------------------------------
\subsection{The Heterogeneous Delays SNN (HD-SNN) model}
%----------------------------%
%
\begin{figure}%[t!]
    \centering
    \includegraphics[width=0.980\linewidth]{figures/HDSNN_graph.pdf}
    \caption{\textbf{The core mechanism of the HD SNN model.} \textit{(Top-Left)}~Two spiking motifs are emitted from four presynaptic neurons. Once integrated by the synapses of the postsynaptic neuron~\textit{(Right)}, the spiking motifs are shifted in time by the synaptic delays and weighted by the synaptic weights~\textit{(Middle-Left)}. When they reach the soma of the postsynaptic neuron, the different spikes contribute to a modification of its membrane potential according to an activation function. In this example, we use the same activation function as for a Leaky Integrate and Fire neuron~\textit{(Bottom-Left)}. The first spiking motif is synchronized by the synaptic delays and causes a sudden rise in the membrane potential of the postsynaptic neuron. An output spike is emitted when the membrane potential reaches the threshold and it is then reset.   \textit{(Right)}~An illustration of a spiking neuron with different synaptic weights (represented by the thickness of the synapses) and different synaptic delays (represented by the length of the synapses). }%
    \label{fig:izhikevich}%
\end{figure}%
% 
%----------------------------%
In a neurobiological recording or in the sensory signal obtained from an event-based camera, the input consists of a stream of \emph{spikes} or \emph{events}. This can be formalized as a list of tuples, each tuple representing neural addresses and timestamps. We defined this list as $\event = \{(\presynaddr_\arank, \timev_\arank)\}_{\arank \in [1,\numevent]}$ where $\numevent \in \mathbb{N} $ is the total number of events in the data stream and the rank $\arank$ is the index of each event in the list of events (\seeFig{izhikevich}-\textit{(Top-Left} for an illustration). Events are usually ordered by their time of occurrence. Each event has a time of occurrence $\timev_\arank$ and an associated address $\presynaddr_\arank$. This defines an address space $\presynaddrspace$, which consists of the set of possible addresses. In a neurobiological recording, this may be the identified set of all presynaptic neurons. For neuromorphic hardware like the output of a DVS, this can be defined as $[1, \Npol] \times [1, \Nx]\times[1, \Ny] \subset \mathbb{N}^3$, where $\Npol$ is the number of polarities ($\Npol=2$ for the ON and OFF polarities encoded in event-based cameras) and $(\Nx, \Ny)$ is the height and width of the image in pixels. Thus, an address $\presynaddr_\arank$ is typically in the form $(\polev_\arank, x_\arank, y_\arank)$ for event-based cameras.

@@

After training our model, we first analyze the weights learned for the different neurons (\seeFig{kernels}). We first notice that these cells are very localized around linear trajectories in the space-delay space, and also show an orientation selectivity perpendicular to motion, similarly to what is found for MT neurons~\citep{deangelis_functional_1999}. Focusing on the positive weights, a strong selectivity along specific motion axes is observed for each of the different kernels. These directions can be easily associated with the direction of motion controlled in the natural images and show a typical anisotropic profile~\citep{kaplan_anisotropic_2013}. The negative weights surround these positive weights, providing an enhanced response. For example, the first kernel shows a strong selectivity for a downward direction of motion. Note that as the delays become larger, two effects can be remarked. First, coefficients become lower which is consistent with the fact that trajectories are defined in a piecewise fashion, such that such decrease provides with an optimal integration with respect to this limited memory. Second, coefficients become less localized compared to the kerenl's spatial profile at short delays, consistent with the average diffusion of information included in the generative model and with the diffusion introduced in motion-based prediction models~\citep{perrinet_motion-based_2012,khoei_flash-lag_2017}. We also observed a strong dependence between the weights reaching the ON polarities and those reaching the OFF polarities. In particular, whenever a weight for a given position and delay is positive for one polarity, it will be negative for the other. This property is due to the way the events are generated and the fact that the luminance cannot increase and decrease at the same time.  Interestingly, the relative organization of the receptive fields in quadrature of phase follows a push-pull organization predicted by~\citet{kremkow_push-pull_2016} to explain neurophysiological results obtained after showing similar natural scenes with synthetic eye movements~\citep{baudot_animation_2013}.
